Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-07-13 analysis/paper/paper.qmd
editing results

To further investigate technological strategies, we incorporated orientation statistics as a complementary approach. This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts. We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were finished in Geomagic Wrap, Rhino8 and R.

@@

We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature.

Higher mean edge angles suggest a more spherical geometry. Ten edge intersections were randomly sampled per spheroid. Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R.

Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface. We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex [@Muller2023]. The deviation between actual surface points and this fitted plane reflects surface roughness. The calculation steps were completed in Python.

Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies. We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes. This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution. Spheroids and sub-spheroids from the QSY site were included in the analysis.

@@

## Results of Morphometric analysis

@@

We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra. The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.

We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20,@fig-SPHARM-Reconstruction-by-Degree). This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon. After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.

A Variance Analysis was performed on the power of each harmonic degree. The results indicate that l= 2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation. Specifically, l=2 reflects global asymmetries—such as elongation or flattening—while degree 3 is more sensitive to localized features like curvature and edge protrusions.

To visualize the structure of shape variation, we applied UMAP to the normalized power spectrum. The UMAP projection reveals distinct clusters in the first two dimensions of the shape space. These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters). This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features. Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples). However, some Multifacial cores and Polyhedron also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.

We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$. The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation (@fig-Correlation-SPHARMpower-Sphericity).

The spherical harmonic energy (SHE), a novel metric introduced herein, quantifies the geometric approximation of core morphology to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity.

Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`). The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids in QSY.

The pursuit of a standardized spherical form among spheroids is further supported by our findings. Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids. @Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts. Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that sub/spheroids maintain a highly standardized rounded shape. In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.

## Results of technological analysis

>>>>

To further investigate technological strategies, we incorporated orientation statistics as a complementary approach. This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts. We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were carried out in Geomagic Wrap, Rhino8 and R.

@@

We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature. Higher mean edge angles suggest a more spherical geometry. Ten edge intersections were randomly sampled per spheroid. Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R. Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface. We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex [@Muller2023]. The deviation between actual surface points and this fitted plane reflects surface roughness. Theses calculations were completed in Python.

Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies. We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes. This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution. Spheroids and sub-spheroids from Qianshangying were included in the analysis.

@@

## Results of the morphometric analysis

@@

We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

Our sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra. The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.

We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree). This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon <!-- BM: please briefly define this --> . After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.

We performed a Variance Analysis on the power of each harmonic degree. The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation. Specifically, l=2 reflects global asymmetries—such as elongation or flattening—while degree 3 is more sensitive to localized features like curvature and edge protrusions. <!-- BM: can we show the results of this somehow, e.g. with a figure? -->

To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum. The UMAP reveals distinct clusters in the first two dimensions of the shape space. These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters). This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features. Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples). However, some Multifacial cores and Polyhedron also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.  <!-- BM: this is an excellent analysis of these results! I'm not entirely clear to me what the difference between these two figures is, I think we might be able to combine them? -->

The UMAP-based groupings also correspond to significant differences in sphericity, as indicated by a Kruskal–Wallis test ($\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$). The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation (@fig-Correlation-SPHARMpower-Sphericity).

The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of core morphology to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity.

Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`). The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying <!-- BM: I think a boxplot would be good to show this to the reader also -->

The pursuit of a standardized and distinct spherical form among spheroids is further supported by our findings. Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids. @Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts. Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that sub/spheroids maintain a highly standardized rounded shape. In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively. <!-- BM: here's another opportunity for a boxplot -->

## Results of the technological analysis
